Viscous Transport in Eroding Porous Media

### 2019-10-03 docs/transport/erosion_transport.tex
NM
finished nick's revisions

\begin{abstract} 
  We extend our previous work, [{\em Journal of Computational Physics},
  {\bf 375}:1--21, 2018], to simulate fluid-mechanical erosion of
  two-dimensional porous media with nearly-touching grains.  The method
  continues to use a boundary integral equation (BIE) method to solve
  the fluid equations, and a stable second-order time stepping method to
  erode the grains.  In our previous work, we required that the eroding
  grains were sufficiently separated to guarantee that the trapezoid
  rule provides a sufficient amount of accuracy. We now remove this
  restriction by introducing a Barycentric quadrature rule whose error
  is bounded independent of the geometry.  The Barycentric quadrature
  rule is computationally expensive, so it is only used when necessary,
  and the spectrally accurate fast-multipole-accelerated trapezoid rule
  is used otherwise. We also simulate and analyze transport through
  eroded porous media.  By using the Barycentric quadrature rule, the
  analysis examines flow both in the fast and slow regions with
  high-order accuracy. Our results reveal several interesting  effects
  of erosion on the tortuosity, dispersion, and pore size distributions.
\end{abstract}

@@

\todo[inline]{Nick made it to here.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Low Porosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a second example with 20 eroding bodies, but with a smaller
initial porosity.  In Figure~\ref{fig:ErodingLow20vort}, we plot the
eroding geometry and vorticity at four evenly spaced instances in time.
Initially, the smallest distance between pairs of bodies is $3.29 \times
10^{-4}$, and the smallest distance between the bodies and solid wall is
$4.50 \times 10^{-3}$.  At these distances, a resolution of
approximately $N_\iin = 27,000$ and $N_\out = 18,000$ discretization
points is required to satisfy the $5h$ threshold \nick{\st{required to guarantee
that} needed for} the trapezoid rule to achieve machine precision.  By applying the
Barycentric quadrature rule, we stably simulate erosion at the much
coarser resolution of $N_\iin = 256$ and $N_\out = 1024$ discretization
points.

@@

To avoid computing streamlines at every time step during the erosion
process, we compute the tortuosity using the Eulerian
method~\eqref{eqn:tortuosity2}.  The initial porosity is $30.67\%$ and
the initial tortuosity is $T = 1.24$.  In
Figure~\ref{fig:ErodingLow20Transport}(a), we plot the tortuosity with
respect to the porosity (blue) and the line of best fit (black) using
the power law 
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.1669}.
\end{align}
This model outperformed the other three models in
equation~\eqref{eqn:tortuosityModels}, and its root-mean-squared error
is $1.13 \times 10^{-2}$.  As grains erode, there is an increase in the
number of streamlines that take a nearly direct \nick{\st{pact} path} through the
geometry, and this decreases the tortuosity. However, the channelization
effect of erosion results in an increase in the tortuosity since the
length of many of the streamlines increases when they deflect from a
high porosity region (low pressure) to a low porosity region (high
pressure). For this example, we see that the net effect is a decrease in
the tortuosity for all time.

In Figure~\ref{fig:ErodingLow20Transport}(b), we plot the temporal
evolution of the particle spreading $\sigma_\lambda$.  As in the last
example, we analyze the spreading at several different porosities and we
use the reinsertion algorithm described in Section~\ref{sec:dispersion}
to form sufficiently long trajectories to observe an asymptotic
dispersion rate.  For all the porosities, the dispersion is much closer
to ballistic when compared to the results in
Figure~\ref{fig:Eroding20Transport}.  However, there are still clear
transitions from ballistic dynamics to asymptotic super-dispersive
spreading.  In contrast to the higher porosity initial condition
(Section~\ref{sec:Eroding20}), at early times the erosion results in a
decrease in the dispersion rate. In particular, \nick{once erosion has eroded}
\todo[inline]{Erosion has eroded?}
the first 5\% of the bodies, the particle spreading transitions from
$\sigma_\lambda \sim t^{0.92}$ to $\sigma_\lambda \sim t^{0.69}$.  To
explain this behavior, recall that anomalous dispersion is caused by
tracers spending time in both the fast and slow regimes.  Since the
initial configuration has a reasonably uniform velocity (see
Figure~\ref{fig:ErodingLow20vort}), albeit a small one, the dispersion
is nearly ballistic. However, as the geometry erodes, the flow becomes
more intermittent, and this results in an increased anomalous dispersion
rate~\cite{dea-leb-den-tar-bol-dav2013}.  Then, as the bodies continue
to erode, the geometry channelizes, and most tracers are transported
with a large velocity through the channels, again resulting in a nearly
ballistic motion~\cite{sie-ili-pri-riv-gua2019}.

@@

We compute the tortuosity using both the
Lagrangian~\eqref{eqn:tortuosity1} and Eulerian
methods~\eqref{eqn:tortuosity2}. Therefore, we must compute the velocity
at $N_p = 1000$ points along the inlet $x=-1$. These normalized
velocities are plotted in Figure~\ref{fig:Eroding100tort}(a) for the
eroded geometry at porosity $\phi = 62.98\%$
(Figure~\ref{fig:Eroding100tort}(c)).  The initial velocity of the
tracers is qualitatively similar to the 20 body example
(Figure~\ref{fig:Eroding20tort}(a)), except with additional maximum and
minimums because of the additional bodies.  Next, in
Figure~\ref{fig:Eroding100tort}(b), we plot the local tortuosity by
finding the length of the streamlines as they pass from $x=-1$ to $x=1$.
Compared to Figure~\ref{fig:Eroding20tort}(b), the local tortuosity is
much more discontinuous.  These discontinuities can be explained by
examining the trajectories of tracers in
Figure~\ref{fig:Eroding100tort}(c).  Here, there are many instances of
nearby streamlines that are deflected apart from one another as they
tend to a stagnation point in the flow, and this results in trajectories
with significantly different lengths.  At this porosity, one of the
tracers travels 25.5\% farther than it would have if the bodies had been
absent, and the average tracer travelled 12\% farther \nick{\st{than it would have
if the bodies were absent,} indicating a tortuosity of $1.12$.}
\todo[inline]{Here and in a previous example, I believe the qualifier `than it would have if the bodies were absent' is obvious from the first part of the sentence and does not need to be repeated because it makes the sentence very verbose.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
As a continuation of our previous work~\cite{qua-moo2018}, we have
simulated dense suspensions and characterized transport in viscous
erosion of porous media. A BIE formulation is used to solve the Stokes
equations and to compute the shear stress. However, to simulate
nearly-touching grains and to consider transport, it is necessary to
integrate near-singular functions. This is achieved by extending a
Barycentric quadrature rule~\cite{bar-wu-vee2015} to accurately compute
the erosion rate and vorticity. By using this quadrature rule, we are
able to form stable simulations of erosion with $N = \mathcal{O}(100)$
discretization points, while the trapezoid rule would require
$\mathcal{O}(10^5)$ discretization points.

The transport is characterized in terms of tortuosity and anomalous
dispersion. The local tortuosity agrees qualitatively with
other works~\cite{mat-kha-koz2008}. However, the tortuosity of the
eroded geometries can not be completely described in terms of the
porosity. In particular, we observe that for certain porous media, the
tortuosity can increase when the porosity increases under erosion. We
also observe super-dispersive spreading, and the rate of dispersion
significantly depends not only on the porosity, but also the number of
eroding bodies.

@@

As a long term goal, we would like to include the effect of inertia and
other transport models. Including inertia requires an integral equation
formulation for the Navier-Stokes equations, and one approach has
recently been developed~\cite{kli-ask-kro2019}. Regarding other
transport models, this would involve a diffusive term to consider the
transport of heat or a contaminant. Forming high-fidelity simulations of
such an advection-diffusion equation can be accomplished by using time
splitting methods and recent work on heat solvers in complex
geometries~\cite{fry-kro-tor2019}.

>>>>

\begin{abstract} 
  We extend our previous work, [{\em Journal of Computational Physics},
  {\bf 375}:1--21, 2018], to simulate fluid-mechanical erosion of
  two-dimensional porous media with nearly-touching grains.  The method
  continues to use a boundary integral equation (BIE) method to solve
  the fluid equations, and a stable second-order time stepping method to
  erode the grains.  In our previous work, we required eroding grains
  that were sufficiently separated to guarantee that the trapezoid rule
  provides a sufficient amount of accuracy. We now remove this
  restriction by introducing a Barycentric quadrature rule whose error
  is bounded independent of the geometry.  The Barycentric quadrature
  rule is computationally expensive, so it is only used when necessary,
  and the spectrally accurate fast-multipole-accelerated trapezoid rule
  is used otherwise. We also simulate and analyze transport through
  eroded porous media.  By using the Barycentric quadrature rule, the
  analysis examines flow both in the fast and slow regions with
  high-order accuracy. Our results reveal several interesting  effects
  of erosion on the tortuosity, dispersion, and pore size distributions.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Low Porosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a second example with 20 eroding bodies, but with a smaller
initial porosity.  In Figure~\ref{fig:ErodingLow20vort}, we plot the
eroding geometry and vorticity at four evenly spaced instances in time.
Initially, the smallest distance between pairs of bodies is $3.29 \times
10^{-4}$, and the smallest distance between the bodies and solid wall is
$4.50 \times 10^{-3}$.  At these distances, a resolution of
approximately $N_\iin = 27,000$ and $N_\out = 18,000$ discretization
points is required to satisfy the $5h$ threshold needed for the
trapezoid rule to achieve machine precision.  By applying the
Barycentric quadrature rule, we stably simulate erosion at the much
coarser resolution of $N_\iin = 256$ and $N_\out = 1024$ discretization
points.

@@

To avoid computing streamlines at every time step during the erosion
process, we compute the tortuosity using the Eulerian
method~\eqref{eqn:tortuosity2}.  The initial porosity is $30.67\%$ and
the initial tortuosity is $T = 1.24$.  In
Figure~\ref{fig:ErodingLow20Transport}(a), we plot the tortuosity with
respect to the porosity (blue) and the line of best fit (black) using
the power law 
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.1669}.
\end{align}
This model outperformed the other three models in
equation~\eqref{eqn:tortuosityModels}, and its root-mean-squared error
is $1.13 \times 10^{-2}$.  As grains erode, there is an increase in the
number of streamlines that take a nearly direct path through the
geometry, and this decreases the tortuosity. However, the channelization
effect of erosion results in an increase in the tortuosity since the
length of many of the streamlines increases when they deflect from a
high porosity region (low pressure) to a low porosity region (high
pressure). For this example, we see that the net effect is a decrease in
the tortuosity for all time.

In Figure~\ref{fig:ErodingLow20Transport}(b), we plot the temporal
evolution of the particle spreading $\sigma_\lambda$.  As in the last
example, we analyze the spreading at several different porosities and we
use the reinsertion algorithm described in Section~\ref{sec:dispersion}
to form sufficiently long trajectories to observe an asymptotic
dispersion rate.  For all the porosities, the dispersion is much closer
to ballistic when compared to the results in
Figure~\ref{fig:Eroding20Transport}.  However, there are still clear
transitions from ballistic dynamics to asymptotic super-dispersive
spreading.  In contrast to the higher porosity initial condition
(Section~\ref{sec:Eroding20}), at early times the erosion results in a
decrease in the dispersion rate. In particular, the first 5\% of the
bodies have eroded, the particle spreading transitions from
$\sigma_\lambda \sim t^{0.92}$ to $\sigma_\lambda \sim t^{0.69}$.  To
explain this behavior, recall that anomalous dispersion is caused by
tracers spending time in both the fast and slow regimes.  Since the
initial configuration has a reasonably uniform velocity (see
Figure~\ref{fig:ErodingLow20vort}), albeit a small one, the dispersion
is nearly ballistic. However, as the geometry erodes, the flow becomes
more intermittent, and this results in an increased anomalous dispersion
rate~\cite{dea-leb-den-tar-bol-dav2013}.  Then, as the bodies continue
to erode, the geometry channelizes, and most tracers are transported
with a large velocity through the channels, again resulting in a nearly
ballistic motion~\cite{sie-ili-pri-riv-gua2019}.

@@

We compute the tortuosity using both the
Lagrangian~\eqref{eqn:tortuosity1} and Eulerian
methods~\eqref{eqn:tortuosity2}. Therefore, we must compute the velocity
at $N_p = 1000$ points along the inlet $x=-1$. These normalized
velocities are plotted in Figure~\ref{fig:Eroding100tort}(a) for the
eroded geometry at porosity $\phi = 62.98\%$
(Figure~\ref{fig:Eroding100tort}(c)).  The initial velocity of the
tracers is qualitatively similar to the 20 body example
(Figure~\ref{fig:Eroding20tort}(a)), except with additional maximum and
minimums because of the additional bodies.  Next, in
Figure~\ref{fig:Eroding100tort}(b), we plot the local tortuosity by
finding the length of the streamlines as they pass from $x=-1$ to $x=1$.
Compared to Figure~\ref{fig:Eroding20tort}(b), the local tortuosity is
much more discontinuous.  These discontinuities can be explained by
examining the trajectories of tracers in
Figure~\ref{fig:Eroding100tort}(c).  Here, there are many instances of
nearby streamlines that are deflected apart from one another as they
tend to a stagnation point in the flow, and this results in trajectories
with significantly different lengths.  At this porosity, one of the
tracers travels 25.5\% farther than it would have if the bodies had been
absent, and the average tracer travelled 12\% farther indicating a
tortuosity of $1.12$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
\label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
As a continuation of our previous work~\cite{qua-moo2018}, we have
simulated dense suspensions and characterized transport in viscous
erosion of porous media. A BIE formulation is used to solve the Stokes
equations and to compute the shear stress. However, to simulate
nearly-touching grains and to consider transport, it is necessary to
integrate near-singular functions. This is achieved by extending a
Barycentric quadrature rule~\cite{bar-wu-vee2015} to accurately compute
the erosion rate and vorticity. By using this quadrature rule, we are
able to perform stable simulations of erosion with $N =
\mathcal{O}(100)$ discretization points, while the trapezoid rule would
require $\mathcal{O}(10^5)$ discretization points.

The transport is characterized in terms of tortuosity and anomalous
dispersion. The local tortuosity agrees qualitatively with other
works~\cite{mat-kha-koz2008}. However, the tortuosity of the eroded
geometries cannot be completely described in terms of the porosity. In
particular, we observe that for certain porous media, the tortuosity can
increase when the porosity increases under erosion. We also observe
super-dispersive spreading, and the rate of dispersion significantly
depends not only on the porosity, but also the number of eroding bodies.

@@

As a long term goal, we would like to include the effect of inertia and
other transport models. Including inertia requires an integral equation
formulation of the Navier-Stokes equations, which is an active area of
research with a few promising algorithms recently
proposed~\cite{gray2019boundary, kli-ask-kro2019}.  Regarding other
transport models, this would involve a diffusive term to consider the
transport of heat or a contaminant. Forming high-fidelity simulations of
such an advection-diffusion equation can be accomplished by using time
splitting methods and recent work on heat solvers in complex
geometries~\cite{fry-kro-tor2019}.
